Tree height and hydraulic traits shape growth responses across droughts in a temperate broadleaf forest

### 2020-07-24 manuscript/response_to_reviews.Rmd

[*issue #92*](https://github.com/SCBI-ForestGEO/McGregor_climate-sensitivity-variation/issues/92)

**To account for potential trends in individual tree growth prior to the drought, which could be influenced by stand dynamics, we tried an intervention time series analysis (ARIMA). Details on the comparison between $Rt$ and $Rt_{ARIMA}$ are presented in the responses to reviewer 4. In short, the ARIMA method yielded broadly similar results, but tended to produce less reasonable estimates when the difference between the two metrics was large, so we decided to retain $Rt$ as the focal metric. While this unquestionably leads to biased estimates for some trees, we would not expect it to bias our analysis; rather, it is one of the reasons that the amount of variation that can be explained by analyses such as these is relatively low. We now include the following content on ARIMA in the manuscript:**

**"Because the $Rt$ metric could be biased by directional pre-drought growth trends, we also tried an intervention time series analysis (ARIMA, [@R-forecast]) that predicted mean drought-year growth based on trends over the preceding ten years and used this value in place of the five-year mean in calculations of resistance ($Rt_{ARIMA}$= observed $BAI$/ predicted $BAI$). The two metrics were strongly correlated (Fig. **S5**). Visual review of the individual tree-ring sequences with the largest discrepancies between these metrics revealed that $Rt$ was less prone to unreasonable estimates than $Rt_{ARIMA}$, so we selected $Rt$ as our focal metric, presenting parallel results for $Rt_{ARIMA}$ in the Supplementary Info."**

@@

73 The “greater root reach” point is almost universally appealed to—cites for these counterarguments would be useful.

**We now address this in the discussion: "Belowground, taller trees would tend to have larger root systems [@enquist_global_2002], but this does not necessarily imply that they have greater access to or reliance on deep soil-water resources that may be critical during drought. While tree size can correlate with the depth of water extraction [@brum_2018_hydrological], the linkage is not consistent. Shorter trees can vary broadly in the depth of water uptake [@stahl_2013_depth], and larger trees may allocate more to abundant shallow roots that are beneficial for taking up water from rainstorms [@meinzer_partitioning_1999]. Moreover, reliance on deep soil-water resources can actually prove a liability during severe and prolonged drought, as these can experience more intense water scarcity relative to non-drought conditions [@chitra-tarak_roots_2018]."**

73 Also, larger trees would necessarily seem to have greater stem storage capacity and thus, if capacitance really is important, more water volume available for buffering transient drought events, so this is another reason to expect larger individuals to be more rather than less resistant.

**We have updated our text to read: "Greater stem capacitance (*i.e.*, water storage capacity) of larger trees may also confer resistance to transient droughts [@phillips_reliance_2003*; @scholz_hydraulic_2011].**

@@

77 In the concluding sentence, “greater growth reductions” came as a surprise because I thought that the focus was greater vulnerability/ mortality. Also, there are convincing theoretical and empirical reasons to think that large trees don’t slow down e.g. (Sillett et al., 2010; Selaya & Anten, 2010; Michaletz et al., 2014).

**We've ensured that language at the start of the paragraph refers to growth: "Many studies have shown that within and across species, large trees tend to be more affected by drought. Greater growth reductions for larger trees were first shown on a global scale by @bennett_larger_2015, and subsequent studies have reinforced this finding [*e.g.*, @hacket-pain_consistent_2016; @gillerot_tree_2020]."**

@@

102 great questions, though within-species comparisons are likely to be most informative here, and your mixed model design is in accord with this; just being cautious because sometimes it’s seductive to regard selection acting between rather than within species

**Thank you.**

105 this sounds great, but was a little unclear. When you say “first, we test…” I thought that you would examine a relationship between microsite and some tree characteristic, e.g. that the largest trees are in moist microsites. But the next sentence didn’t refer to a tree-microsite correlation but vulnerability-size. It would be good to clarify the prose here because these are very important passages.  The third sentence did seem to be about tree-microsite correlations.

**We have considerably revised this paragraph. It now reads as follows: "...First, we focus on how tree size, alone and in its interaction with microenvironmental gradients, influences drought resistance. We examine the contemporary relationship between tree height and microenvironment, including growing season meteorological conditions and crown exposure. We then test whether, consistent with most forests globally, larger-diameter, taller trees tend to have lower drought resistance in this forest, which is in a region (eastern North America) represented by only two studies in the global review of Bennett et al. (2015). We also test for an influence of potential access to available soil water, which should be greater for larger trees in dry but not in perpetually wet microsites. Finally, we focus on the role of species' traits, testing the hypothesis that species’ traits-–particularly leaf drought tolerance traits-–predict drought resistance. We test predictions that drought resistance is higher in ring-porous than semi-ring and diffuse-porous species and that it is correlated with wood density--either positively [@greenwood_tree_2017] or negatively [@hoffmann_hydraulic_2011] and positively correlated with $LMA$. We further test predictions that species with low $PLA_{dry}$ have higher drought resistance, and that species whose leaves lose turgor lower water potentials (more negative $\pi_{tlp}$) have higher resistance."**

@@

112 wasn’t clear where the “ring porous more drought resistant” hypothesis came from. Global data show that ring porous species are almost entirely restricted to the north temperate zone, especially deciduous broadleaf forests (Wheeler et al., 2007). They are virtually absent in drylands, especially warm tropical drylands. Species in drylands are almost overwhelmingly diffuse porous; the high wood density species in chaparrals and in some desert plants can be semi ring porous (Olson et al., in press).

**This is a great point. The hypothesis refers to temperate deciduous biome, which we now clarify in the sentence in question:**

**"Studies in temperate broadleaf forests have observed that ring-porous species showing higher drought tolerance than diffuse-porous species [@friedrichs_species-specific_2009; @elliott_forest_2015; @kannenberg_linking_2019], but this distinction would not hold in the global context [@wheeler_variations_2007, @olson_xylem_2020] and does not resolve differences among the many species within each category."**

### 2020-07-24 manuscript/climate_sensitivity_draft.Rmd
revamped last paragraph
#120

Forests play a critical global role in climate regulation [@bonan_forests_2008], yet there remains enormous uncertainty as to how the forest-dominated terrestrial carbon sink will respond to climate change [@friedlingstein_climatecarbon_2006]. An important aspect of this uncertainty lies with physiological responses of trees to drought [@kennedy_implementing_2019]. In many forested regions around the world, the risk of severe drought is increasing [@trenberth_global_2014; @dai_climate_2018], often despite increasing precipitation [@intergovernmental_panel_on_climate_change_climate_2015;  @cook_unprecedented_2015]. Droughts, intensified by climate change, have been affecting forests worldwide and are expected to continue as one of the most important drivers of forest change in the future [@allen_global_2010; @allen_underestimation_2015]. Understanding forest responses to drought requires elucidation of how tree size, microenvironment, and species’ traits jointly influence individual-level drought resistance, defined here as a tree's ability to maintain growth during drought, and the extent to which their influence is consistent across droughts. 
Because the resistance and resilience of growth to drought is linked to trees' probability of surviving drought [@desoto_low_2020; @liu_hydraulic_2019], understanding growth responses can also help elucidate which trees are most vulnerable to drought-induced mortality.
However, it has proven difficult to resolve the many factors affecting tree growth during drought with available forest census data, which only rarely captures extreme drought, and with tree-ring records, which capture multiple droughts but usually only sample a subset of a forest community, typically focusing on a single species or the largest individuals.

@@

As climate change drives increasing drought in many of the world's forests [@trenberth_global_2014; @intergovernmental_panel_on_climate_change_climate_2015], the fate of forests and their climate feedbacks will be shaped by the biophysical and physiological drivers observed here. 
Our results, consistent with other observations around the world, imply that the tallest, most exposed trees will be most affected [@bennett_larger_2015; @stovall_tree_2019]. 
We show that, at least within the mature forest studied here, the vulnerability conferred by tall height and associated crown exposure outweigh any advantage of a larger root system, even in drier microenvironments. 
This would suggest that the drought responses of trees in mature forests are more strongly differentiated along the size spectrum by their above- than below-ground environment. 
The same may not be true of systems where short trees exist outside of a buffered understory environment--*i.e.*, open grown trees or short-statured, early-successional forests.
The latter appear to be limited more strongly by root water access during drought [@bretfeld_plant_2018], and would also be dominated by species with different traits. 
The earlier-successional species at our site (*Liriodendron tulipifera*, *Quercus spp.*, *Fraxinus americana*) display a mix of traits conferring drought resistance (Table 2), while the late-successional *Fagus grandifolia* displayed high drought resistance, in part because it exists primarily within a buffered microenvironment. 
Further research on how leaf drought tolerance traits and drought vulnerability change over the course of succession would be valuable for addressing how drought resistance changes as forests age [e.g. @rodriguez-caton_long-term_2015]. 
In the meantime, the results of this study advance our knowledge of the factors conferring drought resistance in a mature forest, opening the door for more accurate forecasting of forest responses to future drought.

>>>>

Forests play a critical global role in climate regulation [@bonan_forests_2008], yet there remains enormous uncertainty as to how the forest-dominated terrestrial carbon sink will respond to climate change [@friedlingstein_climatecarbon_2006]. An important aspect of this uncertainty lies with physiological responses of trees to drought [@kennedy_implementing_2019]. In many forested regions around the world, the risk of severe drought is increasing [@trenberth_global_2014; @dai_climate_2018], often despite increasing precipitation [@intergovernmental_panel_on_climate_change_climate_2015;  @cook_unprecedented_2015]. Droughts, intensified by climate change, have been affecting forests worldwide and are expected to continue as one of the most important drivers of forest change in the future [@allen_global_2010; @allen_underestimation_2015]. Understanding forest responses to drought requires elucidation of how tree size, microenvironment, and species’ traits jointly influence individual-level drought resistance, defined here as a tree's ability to maintain growth during drought, and the extent to which their influence is consistent across droughts. 
Because the resistance and resilience (i.e., post-drought recovery) of growth to drought are linked to trees' probability of surviving drought [@desoto_low_2020; @liu_hydraulic_2019], understanding growth responses can also help elucidate which trees are most vulnerable to drought-induced mortality.
However, it has proven difficult to resolve the many factors affecting tree growth during drought with available forest census data, which only rarely captures extreme drought, and with tree-ring records, which capture multiple droughts but usually only sample a subset of a forest community, typically focusing on a single species or the largest individuals.

@@

As climate change drives increasing drought in many of the world's forests [@trenberth_global_2014; @intergovernmental_panel_on_climate_change_climate_2015], the fate of forests and their climate feedbacks will be shaped by the biophysical and physiological drivers observed here. 
Our results show that taller, more exposed trees and species with less drought tolerant leaf traits will be most affected, at least in terms of growth during the drought year. 
Resilience and survival are both linked to resistance (**REFS**), implying that the same factors may influence these. 
Indeed, while the influence of $PLA_{dry}$ and $\pi_{tlp}$ on drought resilience and survival remains to be tested, taller trees have lower resilience [@trugman_tree_2018; @gillerot_tree_2020] and survival [@bennett_larger_2015; @stovall_tree_2019]. 
As climate change-driven droughts affect forests worldwide, there is likely to be a shift from mature forests with tall, buffering trees to forests with a shorter overall stature [@mcdowell_pervasive_2020]. 
At this point, species whose drought resistance relies in part on existence within a buffered microenvironment (e.g., *F. grandifolia*) could in turn become more susceptible. 
Here, the relative importance of tree height *per se* versus crown exposure becomes critical, shaping whether the dominant trees of shorter canopies are significantly more drought resistant because of their shorter stature, or whether high exposure makes them as vulnerable as the taller trees of the former canopy.
Studies disentangling the influence of height and exposure on drought tolerance will be critical to answering this question. 
Ultimately, distributions of tree heights and drought tolerance traits across broad moisture gradients suggest that forests exposed to more drought will ultimately shift towards shorter stature (**REFS**) and be dominated by species with more drought-tolerant traits (**REFS**).
Our study helps to elucidate the mechanisms behind these patterns, opening the door for more accurate forecasting of forest responses to future drought.
